\textit{MSF: Modulated Sub-graph Finder}

### 2019-03-04 MSF_Paper_FA.tex
Fabian suggestions added

\usepackage{subcaption}
\usepackage{hyperref}
\usepackage{pdfpages}
\usepackage[normalem]{ulem}
\newcommand{\FA}[1]{\begingroup\color{magenta}#1\endgroup}
\newcommand{\ILH}[1]{\begingroup\color{blue}#1\endgroup}
\newcommand{\TODO}[1]{\begingroup\color{red}#1\endgroup}
%opening
\begin{document}

@@

	Unlike ORA, FCS has no artificial cut-off to define a DEG list. FCS works in
	three steps. First it calculates the gene-level statistics including
	correlation of molecular measurements using differential expression of
	individual genes, ANOVA, t-test and Z-score. In the second step the
	statistics of individual genes in a pathway are transformed to an
	individual pathway-level statistic commonly using Kolmogorov-Smirnov
	statistic, mean or median. Finally the statistical significance of the
	pathway-level statistics is assessed. Although FCS covers some of the
	limitations from ORA, it still ignores the topology of genes in a pathway,
	cross-talk and overlap of the pathways~\cite{Khatri2012,Campos}. Pathway
	topology based methods are similar to FCS except that they consider the
	topology of each gene during the gene-level statistics but still lacks to aim
	to link the different functional pathways~\cite{Khatri2012}. \FA{mention jActiveModules somewhere in the intro}

	On these grounds we propose a novel approach to make use of the rich gene
	and protein interaction annotation resources available, to gain additional
	functional insights from basic DGEA. To this, we start with the
	presupposition that expression of neighboring genes within a functional
	pathway are not independent from each other. Rather, they are often
	regulating each others expression or are part of the same
	regulon~\cite{Michalak}. We understand that the categorization of links
	between genes into labeled pathways is often an arbitrary one, given the
	extensive cross talk between different pathways. Although these categories
	have proven to be useful in many situations, they force a certain
	perspective onto the interpretation of novel data. Based on these two
	principles, we aim to find sub-graphs of connected genes within cell
	signaling network, which exhibit as a whole significant differential
	expression changes. This approach differs in two main aspects from common
	pathway analysis tools. First, it does not aim to identify functional
	pathways enriched in differentially expressed genes, but detects sub-graphs
	or branches in a network graph (potentially spanning more than one
	functionally grouped pathway) which is coherently modulated. Second, it
	aims to improve the DGEA on the gene level, by collecting the information
	of neighboring genes, which as a whole might exhibit prominent enough
	signal to be called; again as a whole, significantly modulated. All of this can be helpful to understand the cause and effect of a stimulus and might inform about potential points of intervention.

	As input, information on functional links between genes provided by
	e.g.~KEGG or Reactome and information on the differential expression status
	of single genes resulting from a DGEA, are required. As a result the
	analysis returns sub-graphs and their joint confidence scores, reflecting
	how the perturbation is migrated through the network. Furthermore, the
	entry points of perturbation in the networks and overlap with conventional
	pathway categories are returned \FA{mention impact score and reliablilty of sources here}. The output is prepared in a directed
	adjacency file, convenient for visualization, e.g., with
	StringApp~\cite{StringApp}, available as a Cytoscape plug-in~\cite{Cyto}.

@@

	In the next step, we check if any of the initial modulated
        sub-graphs could further be extended \FA{beyond the immediate
          neighbouring genes. Instead of testing single genes and
          their compatibility to be added, groups of genes are
          considered.}

        \sout{This is done when the sub-graph could not be extended any more
        to the direct neighboring gene due to a high \textit{p}-value,
        but the neighboring gene to the direct neighboring gene has a
        low \textit{p}-value. This was required to connect together
        low \textit{p}-value gene but incorporating a single high
        \textit{p}-value gene.} If the combined \textit{p}-value of the
        initial modulated sub-graph and the extension genes is smaller
        than the \textit{p}-value of the initial sub-graph the
        extension is accepted. All possible extension paths up to
        \emph{3} (default 2) genes at all nodes in the sub-graph are
        tested. Again, this step is iteratively repeated until no
        further genes are added to the significant differentially
        expressed sub-graphs. This step bridges small gaps of genes
        without a clear differential signal in the DGEA.\newline

@@

        \FA{After detection and extension of the modulated sub-graphs,
          each pair of so far identified sub-graphs is tested if its
          combination scores better than each on its own.}
        \sout{After detection and extension of the modulated
          sub-graphs, they are tested if combining the different
          sub-graphs score better than on their own.} The merging of
        the sub-graphs is done by depth first search traversal from
        one sub-graph to the other sub-graph. If the two sub-graphs
        merge with the connector of at most \emph{3} genes (default 2
        gene) and the combined \textit{p}-value of the merged
        sub-graph including the bridging genes in between is less than
        the individual \textit{p}-values of the two sub-graphs, the
        two sub-graphs are merged together to one big\FA{ger}
        modulated sub-graph. This step is repeated iteratively until
        no sub-graphs \sout{could be merged}\FA{can be merged
          anymore}.\newline

@@

	In a last post processing step \texttt{MSF} identifies the
        trigger points of the modulated sub-graphs. These trigger
        genes are the sources of the sub-graphs with only outgoing
        edges. These genes can be interpreted as the possible entry
        points of perturbation from where the stimulus causes
        downstream effects. Each individual source is given an impact
        score, \FA{expressing the relative number of downstream genes
          within the corresponding sub-graph directly connected by
          directed links. This score can be interpreted as an upper
          limit of how much of the sub-graph's perturbation could have
          been introduced by the respective source, and thereby could
          be helpful to prioritize different identified sources for
          larger sub-graphs.}\sout{depending on the number of down
          stream genes effected in the corresponding sub-graph,
          equipping with a way to choose the most significant genes
          for unknown phenotypes or biological processes.} In the same
        spirit \FA{as sources}, the most downstream genes of the
        modulated sub-graphs are identified and defined as sinks. Due
        to loops not all sub-graphs are guaranteed to have sources or
        sinks.

        \TODO{?what about source reliability?}

@@

	\texttt{MSF} generates a directed network file as an output,
        containing complete directed interactions of all modulated
        sub-graphs identified. This file could be imported into
        Cytoscape~\cite{Cyto} for visualization. \sout{Second output
          is the source \& sink file. It contains} \FA{Additionally, a
          file containing details on all sources and sinks for all
          modulated sub-graphs is reported. Furthermore, for
          facilitated visualisation in Cytoscape, a node attribute
          file is provided, containing the source weightage and the
          log-fold chances of all considered genes}. \sout{The third output,
        source weightage is provided as a node attribute file. It
        could be imported into Cytoscape to see the impact score of
        sources as the size of the node and log-fold change could be
        used to see how the regulation of different gene hubs in the
        modulated sub-graphs look.}

@@

	MSF requires Java version 8 and JDK 1.8. The few package
        dependencies are already been added to the release. MSF runs
        fast on a standard laptop computer and so it has normal system
        requirements \TODO{What is fast and what is normal?}. To run
        MSF, the user must provide two text files, one containing the
        DGEA \TODO{DESeq2 edgR?} and the second one containing
        directed interactions in an adjacency format file. Example
        files and a detailed tutorial to use MSF has been provided on
        github https://github. com/Modulated-Subgraph-Finder/MSF.

@@

	For the earliest time point at 6~hpi, three large modulated
        sub-graphs were identified with 42, 139, and 69 genes.  The
        modulated sub-graphs consist predominantly of cytokines,
        chemokines (CXCL10, CCL8, CXCL9, CXCL11, CXCR4, CCR7, CCL4L1,
        CCL3L1, CCL4, CCL8, CCL20, CCL3, CCL19) and Interleukin genes
        (IL6, IL27, IL23).  IFNB1 and IFNA1 were both identified as
        two of the possible sources in the most significantly
        modulated sub-graph identified with 42 genes. The impact score
        of IFNB1 is 14.5\% and IFNA1 is 8.7\% for the sub-graph they
        belong to.  Most of the sources identified by \texttt{MSF}
        were type~I interferon induced genes (Supplement material).
        At 24~hpi seven modulated sub-graphs were identified with four
        main sub-graphs consisting of 61, 222, 130 and 242 genes,
        others being smaller than 6 genes. Again, IFNB1 and IFNA1 were
        identified as two sources out of the total sources with 3.9\%
        and 1.6\% impact score. For the last time-point 48~hpi, six
        modulated sub-graphs were identified. Three of the sub-graphs
        were less than ten genes and main sub-graphs had 217, 224 and
        276 genes. IFNB1 and IFNA1 were identified as sources in the
        most significantly modulated sub-graph with an impact score of
        2.8\% and 3.7\% (Supplement material). \TODO{was the mentioned
          impact score the highest of all sources? How many sources
          are there?}

@@

	Table~\ref{tab:rawVsHartung} compares the results of
        \texttt{MSF}, namely the number of detected sub-modules and
        their total genes numbers, to a simple analysis of mapping
        significantly modulated genes from the DGEA to the network and
        joining neighbors to modules. The numbers indicate that
        \texttt{MSF} detects less but larger and meaningful \TODO{?}
        sub-modules, applying its statistical test. Furthermore, the
        dependency of the results from the \textit{p}-value cutoff
        choice is demonstrated for the DGEA, which is avoided for
        \texttt{MSF} altogether. It showcases how applying different
        cut-offs to the \textit{p}-value of genes from edgeR to the
        sub-graphs identified by \texttt{MSF} breaks the larger
        sub-graphs to many smaller unconnected sub-graphs, many of
        which are single genes.

@@

	Three main modulated sub-graphs identified by \texttt{MSF} at
        6~hpi are shown in Fig.~\ref{fig:Sub-graph6hpi}. The graphs
        represent the immediate output of the \texttt{MSF}-analysis,
        visualized by StringApp~\cite{StringApp} in
        Cytoscape~\cite{Cyto}. Each node represents a gene part of a
        modulated sub-graph, whereby the associated colors code the
        functional annotation deduced from KEGG Pathways. The
        cross-talk between the pathways and also the multiple
        employment of many genes is evident. The flow of information
        between the sensors and effectors can be perceived given the
        directionality of each interaction, indicated by arrows
        \sout{in this particular example could easily be deduced}.  In
        more detail, sub-graph~1 (bottom) shows how the activation of
        toll-like receptor, cytokine, chemokine activating jak-stat
        and mapk genes, together with TNF leads into apoptosis. The
        next significant sub-graph (sub-graph~2: top right) reveals
        how information from the Extra-cellular matrix (ECM) receptor,
        which are reported to interact with Ebola glycoprotein
        (GP)~\cite{Veljkovic}, chemokines, cytokines, and cytosolic
        DNA sensing are directly or indirectly controlling cell
        growth, differentiation, proliferation and apoptosis. It
        suggests that dysregulation of these pathways is responsible
        for modulation of apoptosis. Eventually, sub-graph~3 (top
        left) demonstrates how INFA1 and INFB1 modulates once more,
        via only a few intermediate steps, the apoptotic response of
        the cell. On the other side cAMP signaling genes activates
        platelet genes.

	This display case might advertise with how little effort
        complex data can be \FA{processed and prepared for
          interpretation by the domain expert, to } \sout{interpreted,
          help to} apprehend the dynamics of the underlying processes
        and suggest testable hypothesis and potential points of
        intervention.

@@

@@

	jActiveModules~\cite{jActiveModules} \TODO{citation missing in
          bib} is a plugin in Cytoscape that searches for molecular
        interaction network to find expression activated
        sub-networks. The method used to score the expression
        activated sub-networks is close to the method used in
        \texttt{MSF}. \TODO{what are the differences?}

	The first time-point of Ebola infection data was analyzed
        using jActiveModule (Version 3.2.1) to compare the modulated
        sub-graphs identified by \texttt{MSF} and jActiveModule. The
        input files were same for both tools, number of modules
        parameter was set to 5 \TODO{what does this parameter
          do?}. jActiveModule identified 5 modules, the module with
        the highest pathScore was selected for comparison with
        \texttt{MSF} identified modulated sub-graphs \TODO{why only 1
          out of 5?}. The module consists of a single graph with 314
        genes. While \texttt{MSF} identified three directed modulated
        sub-graphs with 42, 69 and 139 genes. The overlap of the
        common genes identified between \texttt{MSF} and jActiveModule
        is shown in Fig~\ref{fig:MSFvsJActiveModule}. \sout{It is
          difficult to state which tool identifies more accurate set
          of genes as modulated, since there is no golden standard
          data example to refer.} Although jActiveModule identifies
        single connected module, \texttt{MSF} provides directionality,
        with the identification of possible perturbation sources of
        the sub-graphs. \TODO{find reasons why jActiveModules gives so
          much bigger moduls by comparing the details of the method!}

@@

	Gene enrichment analysis was performed using Reactome analyze
        data tool~\cite{Reactome} (version 67) \TODO{on which
          data?}. Reactome's over-representation analysis tool tests
        whether certain Reactome pathways are enriched for the list of
        genes submitted to it. Genes from \texttt{MSF} identified
        sub-graphs for each time-point were analyzed for gene
        enrichment using this tool.  For comparison the DEG results
        from edgeR for the three time-points were filtered using the
        cut-off of adjusted \textit{p}-value \sout{>}\FA{<} 0.05. This
        DEG list was used for gene enrichment analysis. The
        compression of \texttt{MSF} identified sub-graphs gene list
        and the DEG list analysis is shown in
        Fig~\ref{fig:msfvsreactome}

	All enriched pathways with a cut-off of \textit{p}-value
        \sout{>}\FA{<} 0.05 0.05 for \texttt{MSF} and DEG list for the
        three time-points were selected. The comparison shows most of
        the pathways known from literature to be dis-regulated by
        Ebola infection are enriched in both the enrichment
        analysis. Toll-Like receptor signaling pathway when interacts
        with EBOV glycoprotein (GP), it triggers the activation of
        cytokines~\cite{Olejnik}. Toll-like receptor pathway is
        expected to be dis-regulated in the early stage of infection,
        this pathway was not identified as significantly dis-regulated
        when \textit{p}-value cut off DEG list was analyzed for
        enrichment. Nine toll-like receptor cascades TLR10, TLR2,
        TLR3, TLR4, TLR5, TLR7/8, TLR9, TLR1:TLR2 and TLR6:TLR2 were
        identified as dis-regulated from gene enrichment analysis of
        \texttt{MSF} identified sub-graph genes, not a single one of
        these cascade was shown to be dis-regulated in pathway
        enrichment analysis from DEG cut-off list. Since \texttt{MSF}
        considers the complete DEG results, even the weak signal at
        the earliest time-point was detected; for example Toll-like
        receptor signaling. While \texttt{MSF} is able to catch weak
        signals, it does not provide information about the functional
        relationships among genes like Reactome tool.

@@

	Gene set enrichment analysis (GSEA)~\cite{Subramanian15545}
        \TODO{reference is missing in bib} is a method to identify
        classes of genes or proteins that are over-represented in a
        large set of genes or proteins. GSEA uses statistical
        approaches to identify significantly enriched or depleted
        groups of genes. The complete DEG list from DGEA of the first
        time-point 6 hpi was analyzed using bioconductor package
        GSEABase (version 1.44.0). GSEA was able to identify
        Toll\FA{-}like recpetor, chemokine signaling pathway,
        cytosolic DNA-sensing pathway, Jak-STAT signaling pathway,
        RIG-I-like receptor signaling pathway and apoptosis as the
        highest ranked pathways. Although GSEA identified the
        important pathways for Ebola infection, it did not show how
        the cross-talk and topology of genes between the pathways
        identifed. \texttt{MSF} and GSEA uses complete DEG list
        without any cut-off, is why pathways important for Ebola
        infection showed up even with weak signal genes in it.

@@

	\begin{figure*}[p]
		\centering
		\fbox{
			\begin{minipage}{13 cm}
				\includegraphics[width=13cm]{MSF_Robustness.png}
				\caption{\FA{Recall rates for genes in
                                  }\sout{Shows the percentage of}
                                  \texttt{MSF} identified sub-graph
                                  \sout{genes recall rates} for the
                                  three different time points of EBOV
                                  infection data for 100
                                  simulation\FA{s} where Poisson
                                  distributed noise was added to the
                                  experimental deduced reads per gene
                                  counts.\label{fig:DEGvsMSF}}
				\label{label}
			\end{minipage}
		}
	\end{figure*}

@@

	\begin{figure*}[p]
		\centering
		\fbox{
			\begin{minipage}{15 cm}
				\includegraphics[width=1.01\linewidth]{ReactomeVsMSF.png}
				\caption{The Upset plot shows the
                                  number of shared pathways between
                                  \texttt{MSF} identified sub-graph
                                  gene list and DEG cut-off list for
                                  the three time-points. All
                                  \sout{the} \FA{XXX} different
                                  toll-like receptor cascades are
                                  \sout{shown to be} in the \FA{set
                                    of}\sout{bar with} 164 shared
                                  pathways only between \texttt{MSF}
                                  at different time-points.}
				\label{fig:msfvsreactome}
			\end{minipage}
		}
	\end{figure*}

>>>>

\usepackage{subcaption}
\usepackage{hyperref}
\usepackage{pdfpages}
\usepackage[normalem]{ulem}
\newcommand{\DONE}[1]{\begingroup\color{blue}#1\endgroup}
\newcommand{\FA}[1]{\begingroup\color{magenta}#1\endgroup}
\newcommand{\ILH}[1]{\begingroup\color{blue}#1\endgroup}
\newcommand{\TODO}[1]{\begingroup\color{red}#1\endgroup}
%opening
\begin{document}

@@

	Unlike ORA, FCS has no artificial cut-off to define a DEG list. FCS works in
	three steps. First it calculates the gene-level statistics including
	correlation of molecular measurements using differential expression of
	individual genes, ANOVA, t-test and Z-score. In the second step the
	statistics of individual genes in a pathway are transformed to an
	individual pathway-level statistic commonly using Kolmogorov-Smirnov
	statistic, mean or median. Finally the statistical significance of the
	pathway-level statistics is assessed. Although FCS covers some of the
	limitations from ORA, it still ignores the topology of genes in a pathway,
	cross-talk and overlap of the pathways~\cite{Khatri2012,Campos}. Pathway
	topology based methods are similar to FCS except that they consider the
	topology of each gene during the gene-level statistics but still lacks to aim
	to link the different functional pathways~\cite{Khatri2012}.

	\DONE{Recently systems biology approaches are also helping illuminate the structural and functional design of a cell~\cite{mitra_carvunis_ramesh_ideker_2013}. Mitra and colleagues discuss the different systems biology approaches to identify and measure the modules of inter-molecular interactions steering a particular biological process. These approaches are classified into four categories, identification of "active modules" with molecular profiles, identification of "conserved modules" across species , identification of "differential modules" with multiple conditions and identification of "composite modules" with interactions of different types. Identification of the active modules is one of the most successful approaches, which overlays the modules or networks onto the molecular profiles (transcriptomics, genomics, proteomics) to identify the "active modules". To find these active modules heuristics solutions like simulated annealing (SA), greedy methods, genetic algorithms (GA), network propagation and co-clustering methods are used~\cite{mitra_carvunis_ramesh_ideker_2013}. JActiveModules~\cite{jActiveModules} has been the first of this kind using simulated annealing to find the active modules.}

	\DONE{The motivation was to find the modulated biological processes cross-talking, for which identifying the modulated networks first with the help of systems biology was necessary.} On these grounds we propose a novel approach to make use of the rich gene
	and protein interaction annotation resources available, to gain additional
	functional insights from basic DGEA. To this, we start with the
	presupposition that expression of neighboring genes within a functional
	pathway are not independent from each other. Rather, they are often
	regulating each others expression or are part of the same
	regulon~\cite{Michalak}. We understand that the categorization of links
	between genes into labeled pathways is often an arbitrary one, given the
	extensive cross talk between different pathways. Although these categories
	have proven to be useful in many situations, they force a certain
	perspective onto the interpretation of novel data. Based on these two
	principles, we aim to find sub-graphs of connected genes within cell
	signaling network, which exhibit as a whole significant differential
	expression changes. This approach differs in two main aspecarg1ts from common
	pathway analysis tools. First, it does not aim to identify functional
	pathways enriched in differentially expressed genes, but detects sub-graphs
	or branches in a network graph (potentially spanning more than one
	functionally grouped pathway) which is coherently modulated. Second, it
	aims to improve the DGEA on the gene level, by collecting the information
	of neighboring genes, which as a whole might exhibit prominent enough
	signal to be called; again as a whole, significantly modulated. All of this can be helpful to understand the cause and effect of a stimulus and might inform about potential points of intervention.

	As input, information on functional links between genes provided by
	e.g.~KEGG or Reactome and information on the differential expression status
	of single genes resulting from a DGEA, are required. As a result the
	analysis returns sub-graphs and their joint confidence scores, reflecting
	how the perturbation is migrated through the network. Furtharg1ermore, the
	entry points of perturbation in the networks and overlap with conventional
	pathway categories are returned. These entry points are given impact scores showing the degree of perturbations introduced by them. The reliability of the entry points is checked using a T-test by investigating the significant difference between the downstream and upstream genes of each entry point. The output is prepared in a directed
	adjacency file, convenient for visualization, e.g., with
	StringApp~\cite{StringApp}, available as a Cytoscape plug-in~\cite{Cyto}.

@@

	In the next step, we check if any of the initial modulated
        sub-graphs could further be extended beyond the immediate
          neighbouring genes. Instead of testing single genes and
          their compatibility to be added, groups of genes are
          considered. If the combined \textit{p}-value of the
        initial modulated sub-graph and the extension genes is smaller
        than the \textit{p}-value of the initial sub-graph the
        extension is accepted. All possible extension paths up to
        \emph{3} (default 2) genes at all nodes in the sub-graph are
        tested. Again, this step is iteratively repeated until no
        further genes are added to the significant differentially
        expressed sub-graphs. This step bridges small gaps of genes
        without a clear differential signal in the DGEA.\newline

@@

        After detection and extension of the modulated sub-graphs,
          each pair of so far identified sub-graphs is tested if its
          combination scores better than each on its own. The merging of
        the sub-graphs is done by depth first search traversal from
        one sub-graph to the other sub-graph. If the two sub-graphs
        merge with the connector of at most \emph{3} genes (default 2
        gene) and the combined \textit{p}-value of the merged
        sub-graph including the bridging genes in between is less than
        the individual \textit{p}-values of the two sub-graphs, the
        two sub-graphs are merged together to one bigger
        modulated sub-graph. This step is repeated iteratively until
        no sub-graphs can be merged
          anymore.\newline

@@

	In a last post processing step \texttt{MSF} identifies the
        trigger points of the modulated sub-graphs. These trigger
        genes are the sources of the sub-graphs with only outgoing
        edges. These genes can be interpreted as the possible entry
        points of perturbation from where the stimulus causes
        downstream effects. Each individual source is given an impact
        score, expressing the relative number of downstream genes
          within the corresponding sub-graph directly connected by
          directed links. This score can be interpreted as an upper
          limit of how much of the sub-graph's perturbation could have
          been introduced by the respective source, and thereby could
          be helpful to prioritize different identified sources for
          larger sub-graphs. In the same
        spirit as sources, the most downstream genes of the
        modulated sub-graphs are identified and defined as sinks. Due
        to loops not all sub-graphs are guaranteed to have sources or
        sinks. The reliability of each source is inspected using a T-test. The significant difference between the two groups, genes downstream the source to the genes upstream of the source is determined. This would help to assess if the source identified is reliable and marks the border between two different regulation regimes.

@@

	\texttt{MSF} generates a directed network file as an output,
        containing complete directed interactions of all modulated
        sub-graphs identified. This file could be imported into
        Cytoscape~\cite{Cyto} for visualization. Additionally, a
          file containing details on all sources and sinks for all
          modulated sub-graphs is reported. Furthermore, for
          facilitated visualisation in Cytoscape, a node attribute
          file is provided, containing the source weightage and the
          log-fold chances of all considered genes.

@@

	\DONE{The only system requirements to run MSF are Java version 8 and JDK 1.8. The few package
		dependencies are already been added to the release. To run
		MSF, the user must provide two text files, one containing the
		DGEA from edgeR~\cite{edgeR} or DESeq2~\cite{love2014moderated}} and  the second one containing
        directed interactions in an adjacency format file. Example
        files and a detailed tutorial to use MSF has been provided on
        github https://github. com/Modulated-Subgraph-Finder/MSF.

@@

	For the earliest time point at 6~hpi, three large modulated
        sub-graphs were identified with 42, 139, and 69 genes.  The
        modulated sub-graphs consist predominantly of cytokines,
        chemokines (CXCL10, CCL8, CXCL9, CXCL11, CXCR4, CCR7, CCL4L1,
        CCL3L1, CCL4, CCL8, CCL20, CCL3, CCL19) and Interleukin genes
        (IL6, IL27, IL23).  IFNB1 and IFNA1 were both identified as
        two of the possible sources in the most significantly
        modulated sub-graph identified with 42 genes. The impact score
        of IFNB1 is 14.5\%, \DONE{the highest impact score and IFNA1 is 8.7\%, in top 5 highest impact scores in the sub-graph they
        	belong to} . Most of the sources identified by \texttt{MSF}
        were type~I interferon induced genes (Supplement material).
        At 24~hpi seven modulated sub-graphs were identified with four
        main sub-graphs consisting of 61, 222, 130 and 242 genes,
        others being smaller than 6 genes. Again, IFNB1 and IFNA1 were
        identified as two sources out of the total sources with 3.9\% 
        and 1.6\% impact score. \DONE{IFNB1 was one of the top 5 impact score sources for the corresponding sub-graph}. For the last time-point 48~hpi, six
        modulated sub-graphs were identified. Three of the sub-graphs
        were less than ten genes and main sub-graphs had 217, 224 and
        276 genes. IFNB1 and IFNA1 were identified as sources in the
        most significantly modulated sub-graph with an impact score of
        2.8\% and 3.7\%, \DONE{but not the highest ranked sources} (Supplement material).

@@

	Table~\ref{tab:rawVsHartung} compares the results of
        \texttt{MSF}, namely the number of detected sub-modules and
        their total genes numbers, to a simple analysis of mapping
        significantly modulated genes from the DGEA to the network and
        joining neighbors to modules. The numbers indicate that
        \texttt{MSF} detects less but larger and interpretable
        sub-modules, applying its statistical test. Furthermore, the
        dependency of the results from the \textit{p}-value cutoff
        choice is demonstrated for the DGEA, which is avoided for
        \texttt{MSF} altogether. It showcases how applying different
        cut-offs to the \textit{p}-value of genes from edgeR to the
        sub-graphs identified by \texttt{MSF} breaks the larger
        sub-graphs to many smaller unconnected sub-graphs, many of
        which are single genes.

@@

	Three main modulated sub-graphs identified by \texttt{MSF} at
        6~hpi are shown in Fig.~\ref{fig:Sub-graph6hpi}. The graphs
        represent the immediate output of the \texttt{MSF}-analysis,
        visualized by StringApp~\cite{StringApp} in
        Cytoscape~\cite{Cyto}. Each node represents a gene part of a
        modulated sub-graph, whereby the associated colors code the
        functional annotation deduced from KEGG Pathways. The
        cross-talk between the pathways and also the multiplhttps://github.com/Modulated-Subgraph-Finder/MSF/tree/master/bine
        employment of many genes is evident. The flow of information
        between the sensors and effectors can be perceived given the
        directionality of each interaction, indicated by arrows.  In
        more detail, sub-graph~1 (bottom) shows how the activation of
        toll-like receptor, cytokine, chemokine activating jak-stat
        and mapk genes, together with TNF leads into apoptosis. The
        next significant sub-graph (sub-graph~2: top right) reveals
        how information from the Extra-cellular matrix (ECM) receptor,
        which are reported to interact with Ebola glycoprotein
        (GP)~\cite{Veljkovic}, chemokines, cytokines, and cytosolic
        DNA sensing are directly or indirectly controlling cell
        growth, differentiation, proliferation and apoptosis. It
        suggests that dysregulation of these pathways is responsible
        for modulation of apoptosis. Eventually, sub-graph~3 (top
        left) demonstrates how INFA1 and INFB1 modulates onchttps://github.com/Modulated-Subgraph-Finder/MSF/tree/master/bine more,
        via only a few intermediate steps, the apoptotic response of
        the cell. On the other side cAMP signaling genes activates
        platelet genes.

	This display case might advertise with how little effort
        complex data can be processed and prepared for
          interpretation by the domain expert, to apprehend the dynamics of the underlying processes
        and suggest testable hypothesis and potential points of
        intervention.

@@

	 \DONE{The purpose of benchmark is mainly to show overall capabilities of \texttt{MSF}. \texttt{MSF} was compared to jActivemodules~\cite{jActiveModules} since they use similar approach to find and score the modules. Comparison of \texttt{MSF} to Reactome~\cite{Reactome} was done because  Reactome is the standard tool used when looking for modulated pathways after DGEA. Since \texttt{MSF} and GSEA~\cite{Subramanian15545} use no cut-off to the DGEA, their results were also compared.}

@@

	jActiveModules~\cite{jActiveModules} is a plugin in Cytoscape that searches for molecular
        interaction network to find expression activated
        sub-networks. The method used to score the expression
        activated sub-networks is close to the method used in
        \texttt{MSF}.  \DONE{The difference is in how these sub-networks are identified. \texttt{MSF} starts building the sub-graphs from one gene, incorporating and combining the \textit{p}-value of the next gene, with the check that the combined \textit{p}-value of new sub-graph should be better than the original. On the other hand jActiveModules first transforms all the gene’s \textit{p}-values to z-scores and tries to find connected sets of genes with unexpectedly high levels of differential expression, in this case high z-scores. The overall score of the sub-network is calculated by combining the z-scores of the genes. Next using their extended simulated annealing method jActiveModules finds the user defined number of highest scoring putative modules by toggling multiple nodes to merge additional components.}

	The first time-point of Ebola infection data was analyzed
        using jActiveModules (Version 3.2.1) to compare the modulated
        sub-graphs identified by \texttt{MSF} and jActiveModules. The
        input files were same for both tools, number of modules
        parameter was set to 5 (default). This would report 5 putative modules even when only one of them is good. Out of the 5 putative modules identified by jActiveModules, the module with the highest pathScore was selected for comparison with
        \texttt{MSF} identified modulated sub-graphs. The module consists of a single graph with 314
        genes. While \texttt{MSF} identified three directed modulated
        sub-graphs with 42, 69 and 139 genes. The overlap of the
        common genes identified between \texttt{MSF} and jActiveModules
        is shown in Fig~\ref{fig:MSFvsJActiveModule}. \DONE{JActiveModules identified a single module, since this module had unexpectedly high levels of differential expression. With the extended simulated annealing method, jActiveModules toggled multiple nodes to merge a number of components with low score into this single high-scoring component making it larger than original.} Although jActiveModules was able to identify
        single connected module, \texttt{MSF} provides directionality,
        with the identification of possible perturbation sources of
        the sub-graphs.

@@

	Gene enrichment analysis was performed using Reactome analyze
        data tool~\cite{Reactome} (version 67) on the different time-points of Ebola infection data. Reactome's over-representation analysis tool tests
        whether certain Reactome pathways are enriched for the list of
        genes submitted to it. Genes from \texttt{MSF} identified
        sub-graphs for each time-point were analyzed for gene
        enrichment using this tool.  For comparison the DEG results
        from edgeR for the three time-points were filtered using the
        cut-off of adjusted \textit{p}-value < 0.05. This
        DEG list was used for gene enrichment analysis. The
        compression of \texttt{MSF} identified sub-graphs gene list
        and the DEG list analysis is shown in
        Fig~\ref{fig:msfvsreactome}

	All enriched pathways with a cut-off of \textit{p}-value
        < 0.05 for \texttt{MSF} and DEG list for the
        three time-points were selected. The comparison shows most of
        the pathways known from literature to be dis-regulated by
        Ebola infection are enriched in both the enrichment
        analysis. Toll-Like receptor signaling pathway when interacts
        with EBOV glycoprotein (GP), it triggers the activation of
        cytokines~\cite{Olejnik}. Toll-like receptor pathway is
        expected to be dis-regulated in the early stage of infection,
        this pathway was not identified as significantly dis-regulated
        when \textit{p}-value cut off DEG list was analyzed for
        enrichment. Nine toll-like receptor cascades TLR10, TLR2,
        TLR3, TLR4, TLR5, TLR7/8, TLR9, TLR1:TLR2 and TLR6:TLR2 were
        identified as dis-regulated from gene enrichment analysis of
        \texttt{MSF} identified sub-graph genes, not a single one of
        these cascade was shown to be dis-regulated in pathway
        enrichment analysis from DEG cut-off list. Since \texttt{MSF}
        considers the complete DEG results, even the weak signal at
        the earliest time-point was detected; for example Toll-like
        receptor signaling. While \texttt{MSF} is able to catch weak
        signals, it does not provide information about the functional
        relationships among genes like Reactome tool.

@@

	Gene set enrichment analysis (GSEA)~\cite{Subramanian15545}
        is a method to identify
        classes of genes or proteins that are over-represented in a
        large set of genes or proteins. GSEA uses statistical
        approaches to identify significantly enriched or depleted
        groups of genes. The complete DEG list from DGEA of the first
        time-point 6 hpi was analyzed using bioconductor package
        GSEABase (version 1.44.0). GSEA was able to identify
        Toll\FA{-}like recpetor, chemokine signaling pathway,
        cytosolic DNA-sensing pathway, Jak-STAT signaling pathway,arg1
        RIG-I-like receptor signaling pathway and apoptosis as the
        highest ranked pathways. Although GSEA identified the
        important pathways for Ebola infection, it did not show how
        the cross-talk and topology of genes between the pathways
        identifed. \texttt{MSF} and GSEA uses complete DEG list
        without any cut-off, is why pathways important for Ebola
        infection showed up even with weak signal genes in it.

@@

	\begin{figure*}[p]
		\centering
		\fbox{
			\begin{minipage}{13 cm}
				\includegraphics[width=13cm]{MSF_Robustness.png}
				\caption{Recall rates for genes in
                                  \texttt{MSF} identified sub-graph for the
                                  three different time points of EBOV
                                  infection data for 100
                                  simulations where Poisson
                                  distributed noise was added to the
                                  experimental deduced reads per gene
                                  counts.\label{fig:DEGvsMSF}}
				\label{label}
			\end{minipage}
		}
	\end{figure*}

@@

	\begin{figure*}[p]
		\centering
		\fbox{
			\begin{minipage}{15 cm}
				\includegraphics[width=1.01\linewidth]{ReactomeVsMSF.png}
				\caption{The Upset plot shows the
                                  number of shared pathways between
                                  \texttt{MSF} identified sub-graph
                                  gene list and DEG cut-off list for
                                  the three time-points. All
                                  10 different
                                  toll-like receptor cascades are
                                   in the set
                                    of 164 shared
                                  pathways only between \texttt{MSF}
                                  at different time-points.}
				\label{fig:msfvsreactome}
			\end{minipage}
		}
	\end{figure*}
